---
title: '<span style="font-size: 50px; font-face: bold">Ideas cascading into keystrokes -- Modelling writing hesitations as Bayesian mixture process</span>'
author:  '<span style="font-size: 25px; font-face: bold">Jens Roeser, Mark Torrance (both Nottingham Trent University), Rianne Conijn (Eindhoven University of Technology), Evgeny Chukharev (Iowa State University)</span>'
output: 
  ioslides_presentation:
    incremental: false
    transition: slower
    widescreen: true
    css: ../misc/slides.css
    self_contained: true
#    logo: ../misc/gfx/ntu.png
bibliography      : ["../../report/references.bib"]
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      eval = TRUE, 
                      message = FALSE, 
                      comment=NA,
                      fig.align = "right")
options("kableExtra.html.bsTable" = T, digits = 2)
options(pillar.print_min = 5, pillar.print_max = 6)
library(tidyverse)
library(knitr)
library(kableExtra)
library(gridExtra)
library(patchwork)
library(ggthemes)
library(polspline)
source("../../scripts/functions.R")
source("../scripts/get_pdens_plot.R")

beta_label <- "Fluent interkey\nintervals (in msecs)"
delta_label <- "Slowdown for hesitant\ninterkey intervals (in secs)"
theta_label <- "Probability of hesitant\ninterkey intervals"


theme_set(theme_few(base_size = 12) +
            theme(strip.background = element_blank(),
                  legend.position = "top",
                  legend.justification = "right",
                  panel.grid = element_blank(),
                  panel.background = element_rect(fill = "transparent"), # bg of the panel
                  plot.background = element_rect(fill = "transparent", color = NA)))
```


##




<div style="float: left; width: 70%;">
\
\

```{r fig.width=7}
target <- str_split_1("These are keystroke intervals", pattern = "")
x <- 1:length(target)
y <- c(900,
       rnorm(9, 200, 50), 500, 
       rnorm(9, 200, 50), 400, 
       rnorm(8, 200, 50))

ggplot(data = NULL,aes(y = y, x = x)) +
  geom_path() +
  geom_point(size = 5, shape = 21) +
  theme_classic(base_size = 14) +
  labs(x= "Produced key", 
       y = "Interkey intervals (in msecs)") +
  scale_x_continuous(breaks = x, labels = target) +
  scale_y_continuous(limits = c(0, 1100)) +
  theme(axis.ticks = element_blank())
```

</div>

<div style="float: right; width: 25%;">
\
\

- Understanding factors that influence interkey intervals requires a theory of how keystrokes are sequenced.
- Which processes are required and how are they coordinated?

</div>

# How do we produce texts?

```{r intro, child = "../sections/introduction.Rmd"}
```

```{r intro, child = "../sections/evidence.Rmd"}
```


# How can we represent these models statistcally?

```{r data, child = "../sections/statisticalmodels.Rmd"}
```


# How can we evaluate these models?

```{r data, child = "../sections/datasets.Rmd"}
```

```{r textloc, child = "../sections/textlocations.Rmd"}
```

```{r modeloverview, child = "../sections/modeloverview.Rmd"}
```

```{r implementation, child = "../sections/implementation.Rmd"}
```


# Which model showed better performance?

```{r modelcompresults, child = "../sections/modelcomparisons.Rmd"}
```

```{r fittodata, child = "../sections/modelfit.Rmd"}
```

# How do text locations affect the mixture process?

```{r locresults, child = "../sections/locationeffects.Rmd"}
```


# Planning text unfolds in parallel to production!

```{r sim, child = "../sections/discussion.Rmd"}
```


# "Maybe mixture models just always fit bettter?" 

```{r sim, child = "../sections/simulation.Rmd"}
```


## Thank you!

We are grateful for all authors who made their available (to us), in particular Nina Vandermeulen, Alessandra Rossetti, and colleagues. This work was supported by 

- National Science Foundation under Grant No. 2016868: "ProWrite: Biometric feedback for improving college students' writing processes." (Iowa State University)
- UKRI ESRC under Grant No. ES/W011832/1: "Can you use it in a sentence?: Establishing how word-production
difficulties shape text formation." (Nottingham Trent University)

Data and analysis scripts written in R and Stan can be found on OSF https://osf.io/z65dw/ [@roeser2024osf].

For a tutorial on Bayesian mixture-model analysis in the context of keystroke data see
https://rpubs.com/jensroes/mixture-models-tutorial.


## References {.smaller}

<style>
slides > slide { overflow: scroll; }
slides > slide:not(.nobackground):after {
  content: '';
}
</style>


