## Serial view: single-process model  {.smaller}


<div style="float: left; width: 40%;">

```{r fig.width=5}
linewidth <- .35
data <- data.frame(x = c(200, 1500))
word_mean <- 800
sent_mean <- 950
sd <- c(40, 40)
word_lbl <- "word initial"
sent_lbl <- "sentence initial"

ggplot(data = data, aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = word_mean, sd = sd[1]), 
                linewidth = linewidth,
                aes(colour = "word initial"), show.legend = F) +
  geom_area(stat = "function", fun = dnorm, aes(fill = word_lbl), alpha = .25, 
            args = list(mean = word_mean, sd = sd[1])) +
  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_y_continuous(limits = c(0, .012)) +
  scale_x_continuous(labels = scales::comma, limits = c(650, 1150)) +
  labs(x = "Interkey intervals [in ms]", 
       colour = "Text location",
       fill = "Text location") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())


```
</div>

<div style="float: right; width: 50%;">
\
\
\
\

$$
\text{iki}_i \sim\text{ } \text{log}\mathcal{N}(\beta, \sigma_{e}^2)
$$

</div>




## Serial view: single-process model  {.smaller}


<div style="float: left; width: 40%;">

```{r fig.width=5}
ggplot(data = data, aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = word_mean, sd = sd[1]), 
                linewidth = linewidth,
                aes(colour = "word initial")) +
  stat_function(fun = dnorm, n = 101, args = list(mean = sent_mean, sd = sd[2]), 
                linewidth = linewidth,
                aes(colour = "sentence initial"), show.legend = F) +
  geom_area(stat = "function", fun = dnorm, aes(fill = word_lbl), alpha = .25, 
            args = list(mean = word_mean, sd = sd[1])) +
  geom_area(stat = "function", fun = dnorm, aes(fill = sent_lbl), alpha = .25, 
            args = list(mean = sent_mean, sd = sd[2])) +
  geom_curve(aes(x = word_mean, xend = sent_mean, y = 0.01, yend = 0.01),
             curvature = -0.2, 
             size = .25,
             arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x=(sent_mean+word_mean)/2, y=.0115, label=paste0("beta[diff]"), size = 4, parse=TRUE) +
  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_y_continuous(limits = c(0, .012)) +
  scale_x_continuous(labels = scales::comma, limits = c(650, 1150)) +
  labs(x = "Interkey intervals [in ms]", 
       colour = "Text location",
       fill = "Text location") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```
</div>

<div style="float: right; width: 50%;">
\
\
\
\

$$
\text{iki}_i \sim\text{ } \text{log}\mathcal{N}(\beta_\text{textlocation[i]}, \sigma_{e}^2)
$$

</div>


## Parallel view can be implemented as two distributions mixture-process [@roeser2021modelling]

- Executing motor movements [100-150 msecs, @conijn2019understanding;@waes2019].
- If upstream processes provide output more slowly, then interkey intervals are determined by time taken to complete upstream processes and not motor movement. 


## Parallel view: mixture of two processes {.smaller}

<div style="float: left; width: 40%;">


```{r fig.width=5}
ggplot(data = data, aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = word_mean, sd = sd[1]), 
                linewidth = linewidth,
                aes(colour = "word initial"), show.legend = F) +
  geom_area(stat = "function", fun = dnorm, aes(fill = word_lbl), alpha = .25, 
            args = list(mean = word_mean, sd = sd[1])) +
  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_y_continuous(limits = c(0, .012)) +
  scale_x_continuous(labels = scales::comma, limits = c(650, 1150)) +
  labs(x = "Interkey intervals [in ms]", 
       colour = "Text location",
       fill = "Text location") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())


```


</div>

<div style="float: right; width: 50%;">
\
\
\
\

$$
\text{iki}_{i} \sim\text{ } \theta \times \text{log}\mathcal{N}(\beta, \sigma_{e}^2) \\
\theta=1
$$


</div>







## Parallel view: mixture of two processes {.smaller}


<div style="float: left; width: 40%;">

```{r fig.width=5}
# function for mixture models
plot_mix_comps <- function(x, mu, sigma, lam) lam * dnorm(x, mu, sigma)

# Parameters
data <- data.frame(x = c(50, 1500))
theta_conj <- .3
theta_simp <- .1
alpha <- 250
delta <- c(250, 500)
alpha2_w <- alpha + delta[1]
alpha2_s <- alpha + delta[2]
sigma <- 40
sigma_diff <- c(30, 60)
sigmae <- sigma
sigmae_p_w <- sigma + sigma_diff[1]
sigmae_p_s <- sigma + sigma_diff[2]

ggplot(data, aes(x)) +
  stat_function(geom = "line", fun = plot_mix_comps,
                aes(colour = word_lbl),
                linewidth = linewidth,
                args = list(alpha, sigmae, lam = 1 - theta_simp)) +
  stat_function(geom = "line", fun = plot_mix_comps,
                aes(colour = word_lbl),
                linewidth = linewidth,
                args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
            args = list(alpha, sigmae, lam = 1 - theta_simp)) +
  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
            args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_x_continuous(labels = scales::comma, 
                     limits = c(50, 1000), 
                     breaks = seq(0, 1500, 200)) +
  labs(x = "Interkey intervals [in ms]", 
       colour = "Text location",
       fill = "Text location") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())

```
</div>

<div style="float: right; width: 50%;">
\
\
\
\

$$
\text{iki}_{i} \sim\text{ } \theta \times \text{log}\mathcal{N}(\beta + \delta, \sigma_{e'}^2) + \\
   (1 - \theta) \times \text{log}\mathcal{N}(\beta, \sigma_{e}^2)
$$

- Fluent typing speed: $\beta$
- Hesitation slowdown (pause duration): $\delta$
- Hesitation probability (pause frequency): $\theta$

</div>


