# Fit to data

Model fit was assessed by testing to what extent data simulated from the model match the observed data. From each model we obtained 50 simulations (predictions). Simulated and observed data are visualised in Figure \ref{fig:prediction}. Data predicted by the mixture model were substantially more similar to the observed data compared to any of the single-distribution models. The worse fit was observed for the single-distribution Gaussian model which predicted a larger number of negative keystroke intervals.

```{r eval = F}
theme_set(theme_few(base_size = 10) +
              theme(strip.background = element_blank(),
                    legend.position = "top",
                    legend.justification = "right",
                    panel.grid = element_blank(),
                    panel.background = element_rect(fill = "transparent"), # bg of the panel
                    plot.background = element_rect(fill = "transparent", color = NA),
                    axis.text.y = element_blank(),
                    axis.ticks.y = element_blank(),
                    axis.title = element_blank()))

file <- list.files("../../stanout", pattern = "sims", full.names = T, recursive = T)
ds <- str_remove_all(file, "\\.|\\/|stanout|all_sims.csv") %>% 
    str_to_upper() %>% 
    str_replace(., "PLANTRA", "PLanTra") %>%
    str_replace(., "GUNNEXP2", "GE2")

# remove the shift model
ds <- ds[!str_detect(ds, "SHIFT")]

# get all density plots
plots <- map(1:length(ds), ~get_pdens_plot(file[.], max_iki = 2000)) 

# remove strips after first row
plots[2:length(ds)] <- map(2:length(ds), ~plots[[.]] + theme(strip.text = element_blank()))

# create labels for datasets
labs <- map(ds, ~ggplot(mapping = aes(x = 0, y = 0, label = .x)) + 
  geom_text(angle = 270, size = 2.5) +
  theme_void())

# add names for data sets on the right
plots_labs <- map(1:length(ds), ~plots[[.]] + labs[[.]] + plot_layout(widths = c(.95, .05)))

# combine legends
plots_ledg <- wrap_plots(plots_labs, guides = "collect", ncol = 1)

# add axe titles
plots_fin <- grid.arrange(patchworkGrob(plots_ledg), 
                            left = "Density", bottom = "Interkey interval in secs")
  
# save plot 
ggsave("../figures/fitplots.pdf", plots_fin, height = 7.25, width = 6.25)

```


```{r prediction, fig.cap="Predicted data (50 simulations in each cell) in red compared to interkey interval data shown in black shown by model and datasets. Better fit i shown on the left and weaker fit is shown on the right. X-axes were truncated at 2 secs.", eval = T, fig.height=8.25, fig.width=6.75}
knitr::include_graphics("figures/fitplots.pdf")
```






