
```{r}
# get posterior
file <- list.files(path = "../../stanout/cato", 
                   pattern = "mogbetaco.+.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob")) 

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n()) 

ps_cato <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>% 
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(task, group, location, name)) %>% 
  mutate(group = str_c("Task: ", task, "; Group: ", group),
         ds = "CATO") %>% 
  select(-task)

file <- list.files(path = "../../stanout/c2l1", 
                   pattern = "mogbetacon.*.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob")) 

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n()) 

ps_c2l1 <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta,by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>% 
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(location, name)) %>% 
  mutate(ds = "C2L1",
         group = "1") 

file <- list.files(path = "../../stanout/gunnexp2", 
                   pattern = "mogbetacon.*.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob")) 

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n()) 

ps_gunn <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(location, xn, name)) %>% 
  mutate(ds = "GE2") %>% 
  rename(group = xn)

file <- list.files(path = "../../stanout/lift", 
                   pattern = "mogbetaco.+.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob")) 

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n()) 

ps_lift <- ps %>% filter(param != "beta") %>% 
  mutate(across(genre, ~ifelse(is.na(genre), "Arg", .))) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(genre, topic, location, name)) %>% 
  mutate(group = str_c("Topic: ", topic, "; Genre: ", genre),
         ds = "LIFT") %>% 
  select(-topic, -genre)

file <- list.files(path = "../../stanout/plantra", 
                   pattern = "mogbetaco.+.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob")) 

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n()) 

ps_plantra <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(task, location, name)) %>% 
  mutate(ds = "PLanTra") %>% 
  rename(group = task)

file1 <- list.files(path = "../../stanout/spl2", 
                    pattern = "mogbetaco.+.csv$",
                    full.names = T)

file2 <- list.files(path = "../../stanout/spl2_shift", 
                    pattern = "mogbetaco.+.csv$",
                    full.names = T)

ps <- map_dfr(c(file1, file2), ~read_csv(.) %>% 
                     mutate(data = .x)) %>% 
  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
         across(data, ~recode(., spl2 = "SPL2 (_^[shift])",
                              spl2_shift = "SPL2 (_^[shift] + C)"))) %>% 
  filter(param %in% c("beta", "delta", "prob"))

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value, data) %>% 
  mutate(idx = 1:n(), .by = data) 

ps_spl2_both <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(delta, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx, data)) %>% 
  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
         beta2 = beta + delta,
         across(c(beta2, beta), ~exp(.)),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  select(-beta2) %>% 
  pivot_longer(c(beta, prob, delta)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper),
                   .names = "{.fn}"),
            .by = c(lang, location, name, data)) %>% 
  mutate(ds = "SPL2",
         group = str_c("Data: ", data, "; Language: ", lang)) %>% 
  select(-lang) 

ps_spl2 <- ps_spl2_both %>% filter(data == "SPL2 (_^[shift])")

# Combine all
ps <- bind_rows(ps_cato, ps_c2l1, ps_gunn, ps_lift, ps_plantra, ps_spl2)
```


# Key-combination effect 

The analysed datasets differ to the extent that interkey intervals at before sentence location did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GE2) scope over both the transition to the shift key and the the transition to the character following the shift key. In other words, the pause before sentences summed across two key intervals in the PLanTra and LIFT data, namely `_^[shift]^C`, where `^` represents a transition between keys, but only one key interval, namely `_^[shift]`, for the remaining datasets. Therefore, longer or more frequent pauses for transitions immediately proceeding a sentence could be explained without reference to linguistic edges. 


Therefore we compared whether any difference can be explain on the basis of the additional keystroke involved in before-sentence transitions. We used the SPL2 data for this comparison and calculated intervals including and excluding the keystroke after shift. Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference for before and within-word key-intervals. The results of this comparison can be found in Table \ref{tab:shiftcellmeans} (only for the data from participants writing in their first language). 

```{r shiftcellmeans, results = 'asis'}
# cell means and data set difference
cellmeans <- ps_spl2_both %>% 
  separate(group, into = c("group", "lang"), sep = "; ") %>%
  filter(str_detect(lang, "L1")) %>% 
  select(-ds, -group, -lang) %>%
  mutate(#across(lang, ~str_remove(., "Language: ")),
         across(location, ~str_replace(., "\\n", " "))) %>% 
  pivot_wider(names_from = name, 
              values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"),
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(location, data), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = data, 
              values_from = est) %>% 
  mutate(across(c(location), ~ifelse(param == "beta", "overall", .))) %>% 
  unique()

file1 <- list.files(path = "../../stanout/spl2", 
                    pattern = "mogbetacon.+.csv$",
                    full.names = T)

file2 <- list.files(path = "../../stanout/spl2_shift", 
                    pattern = "mogbetacon.+.csv$",
                    full.names = T)

tmp <- map_dfr(c(file1, file2), ~read_csv(.x) %>% 
                 mutate(data = .x)) %>% 
  filter(param %in% c("beta", "delta", "theta")) %>% 
  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
  filter(lang %in% c("EN", "overall")) %>% 
  mutate(#across(lang, recode, "EN" = "L1", "ES" = "L2"),
         across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .))) %>% 
  pivot_wider(names_from = data, values_from = value) %>% 
  unnest(c(spl2, spl2_shift)) %>% 
  mutate(diff = spl2_shift - spl2) %>% 
  summarise(across(diff, list(mean = mean, 
                              lower = lower, 
                              upper = upper,
                              BF = BF), 
                   .names = "{.fn}"),
            .by = c(location, param)) %>% 
  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
         across(param, recode, theta = "prob")) %>% 
  left_join(cellmeans) %>% 
  filter(location %in% c("overall", "before sentence")) %>% 
  mutate(across(param, ~recode_factor(., 
                                `beta` = "Fluent interkey intervals",
                                `delta` = "Hesitation slowdown",
                                `prob` = "Hesitation probability",
                                .ordered = T))) %>% 
  select(`Model parameter` = param, 
         `_^[shift]^C` = ends_with(" + C)"),
         `_^[shift]` = ends_with("FT])"),
         "Difference" = mean, BF) 


write_csv(tmp, "../tables/shift_effect_cellmeans_constr.csv")

apa_table(tmp,
          align = c("l", "l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for interkey interval immediately preceding a sentence. Cell means are shown for interkey intervals that do and do not involve the interkey intervals to the character following shift in msecs for the slowdown and the probability of hesitant interkey intervals. The difference for including the interkey intervals to the character after shift is shown on log scale (for durations) and logit scale for probability of hesitant interkey intervals; 95% PIs are shown in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          font_size = "footnotesize",
          longtable = T)
          # stub_indents = list("Fluent interkey intervals" = 1,
          #                     "Hesitation duration" = 2:3,
          #                     "Hesitation probability" = 4:5) )


```

The duration of fluent and hesitant interkey intervals was not affected by whether or not the sentence-initial transition include the character following shift. However, we found strong evidence for an increased hesitation probability when the before-sentence key transition included the character following the shift key. Notably, the hesitation probability increased closes to ceiling when the transition following the shift key was added. In other words the mixture model identifies almost all before-sentence transition as being hesitant as they always include two keystrokes while interkey intervals from all other transition location -- the majority of the data -- do not. 