
```{r}
# get posterior
file <- list.files(path = "../../stanout/cato", 
                   pattern = "mogbetaco.+.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob")) 

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n()) 

ps_cato <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
#         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  pivot_longer(c(beta, prob, beta2)) %>% 
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(task, group, location, name)) %>% 
  mutate(group = str_c("Task: ", task, "; Group: ", group),
         ds = "CATO") %>% 
  select(-task)

file <- list.files(path = "../../stanout/c2l1", 
                   pattern = "mogbetacon.*.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob")) 

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n()) 

ps_c2l1 <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta,by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
#         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  pivot_longer(c(beta, prob, beta2)) %>% 
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(location, name)) %>% 
  mutate(ds = "C2L1",
         group = "1") 

file <- list.files(path = "../../stanout/gunnexp2", 
                   pattern = "mogbetacon.*.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob")) 

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n()) 

ps_gunn <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
#         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  pivot_longer(c(beta, prob, beta2)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(location, xn, name)) %>% 
  mutate(ds = "GE2") %>% 
  rename(group = xn)

file <- list.files(path = "../../stanout/lift", 
                   pattern = "mogbetaco.+.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob")) 

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n()) 

ps_lift <- ps %>% filter(param != "beta") %>% 
  mutate(across(genre, ~ifelse(is.na(genre), "Arg", .))) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
#         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  pivot_longer(c(beta, prob, beta2)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(genre, topic, location, name)) %>% 
  mutate(group = str_c("Topic: ", topic, "; Genre: ", genre),
         ds = "LIFT") %>% 
  select(-topic, -genre)

file <- list.files(path = "../../stanout/plantra", 
                   pattern = "mogbetaco.+.csv$",
                   full.names = T)

ps <- read_csv(file) %>% filter(param %in% c("beta", "beta2", "prob")) 

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value) %>% 
  mutate(idx = 1:n()) 

ps_plantra <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta2, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx)) %>% 
  mutate(across(c(beta2, beta), exp),
#         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  pivot_longer(c(beta, prob, beta2)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"),
            .by = c(task, location, name)) %>% 
  mutate(ds = "PLanTra") %>% 
  rename(group = task)

file1 <- list.files(path = "../../stanout/spl2", 
                    pattern = "mogbetaco.+.csv$",
                    full.names = T)

file2 <- list.files(path = "../../stanout/spl2_shift", 
                    pattern = "mogbetaco.+.csv$",
                    full.names = T)

ps <- map_dfr(c(file1, file2), ~read_csv(.) %>% 
                     mutate(data = .x)) %>% 
  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
         across(data, ~recode(., spl2 = "SPL2 (_^[shift])",
                              spl2_shift = "SPL2 (_^[shift] + C)"))) %>% 
  filter(param %in% c("beta", "delta", "prob"))

ps_beta <- filter(ps, param == "beta") %>% 
  select(beta = value, data) %>% 
  mutate(idx = 1:n(), .by = data) 

ps_spl2_both <- ps %>% filter(param != "beta") %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(delta, prob)) %>% 
  mutate(idx = 1:n(), .by = where(is.character)) %>% 
  left_join(ps_beta, by = join_by(idx, data)) %>% 
  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
         beta2 = beta + delta,
         across(c(beta2, beta), ~exp(.)),
#         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  pivot_longer(c(beta, prob, beta2)) %>%
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper),
                   .names = "{.fn}"),
            .by = c(lang, location, name, data)) %>% 
  mutate(ds = "SPL2",
         group = str_c("Data: ", data, "; Language: ", lang)) %>% 
  select(-lang) 

ps_spl2 <- ps_spl2_both %>% filter(data == "SPL2 (_^[shift])")

# Combine all
ps <- bind_rows(ps_cato, ps_c2l1, ps_gunn, ps_lift, ps_plantra, ps_spl2)
```


# Key-combination effect 

Interkey intervals at before sentence location either did (PLanTra, LIFT) scope over both the shift key and the character following the shift key or included only the first key following a the space immediately proceeding a sentence (CATO, C2L1, SPL2, GE2). In other words, the interkey interval before sentences are sum of two interkey intervals in the PLanTra and LIFT data, namely `_^[shift]^C`, where `^` represents a transition between keys, consists only of one interkey interval, namely `_^[shift]`, in the remaining datasets. Therefore, differences in hesitation patterns across datasets for transitions immediately proceeding a sentence could be explained without reference to edges of linguistic units. 

Therefore we tested to what extent the inclusion of one additional interkey interval at sentence boundaries affected the hesitation results. We used the SPL2 data for this comparison and calculated intervals including and excluding the keystroke following a sentence-initial shift press. Although we modelled all transition locations, we present only before-sentence transitions as there was, as one would expect, no difference for before and within-word interkey intervals. The results of this comparison can be found in Table \ref{tab:shiftcellmeans} (only for the data from participants writing in their first language). 

```{r shiftcellmeans, results = 'asis'}
# cell means and data set difference
cellmeans <- ps_spl2_both %>% 
  mutate(across(name, ~recode(., beta2 = "delta"))) %>% 
  separate(group, into = c("group", "lang"), sep = "; ") %>%
  filter(str_detect(lang, "L1")) %>% 
  select(-ds, -group, -lang) %>%
  mutate(#across(lang, ~str_remove(., "Language: ")),
         across(location, ~str_replace(., "\\n", " "))) %>% 
  pivot_wider(names_from = name, 
              values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"),
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(location, data), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = data, 
              values_from = est) %>% 
  mutate(across(c(location), ~ifelse(param == "beta", "overall", .))) %>% 
  unique()

file1 <- list.files(path = "../../stanout/spl2", 
                    pattern = "mogbetacon.+.csv$",
                    full.names = T)

file2 <- list.files(path = "../../stanout/spl2_shift", 
                    pattern = "mogbetacon.+.csv$",
                    full.names = T)

tmp <- map_dfr(c(file1, file2), ~read_csv(.x) %>% 
                 mutate(data = .x)) %>% 
  filter(param %in% c("beta", "delta", "theta")) %>% 
  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
  filter(lang %in% c("EN", "overall")) %>% 
  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .))) %>% 
  pivot_wider(names_from = data, values_from = value) %>% 
  unnest(c(spl2, spl2_shift)) %>% 
  mutate(diff = spl2_shift - spl2) %>% 
  summarise(across(diff, list(mean = mean, 
                              lower = lower, 
                              upper = upper,
                              BF = BF), 
                   .names = "{.fn}"),
            .by = c(location, param)) %>% 
  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
         across(param, recode, theta = "prob")) %>% 
  left_join(cellmeans) %>% 
  filter(location %in% c("overall", "before sentence")) %>% 
  mutate(across(param, ~recode_factor(., 
                                `beta` = "Fluent interkey intervals",
                                `delta` = "Hesitation duration",
                                `prob` = "Hesitation probability",
                                .ordered = T))) %>% 
  select(`Model parameter` = param, 
         `_^[shift]^C` = ends_with(" + C)"),
         `_^[shift]` = ends_with("FT])"),
         "Difference" = mean, BF) 


write_csv(tmp, "../tables/shift_effect_cellmeans_constr.csv")

apa_table(tmp,
          align = c("l", "l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for interkey interval immediately preceding a sentence. Cellmeans are shown for interkey intervals that do and do not include the interkey interval following a sentence-initial shift key press. Hesitation duration is shown in mseecs along with the hesitation probability. The difference due to the additional keystoke is shown fo the hesitation duration (log scale) and hesitation probability (logit scale); 95% PIs are shown in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          font_size = "footnotesize",
          longtable = T)

```

The duration of fluent and hesitant interkey intervals was not affected by whether or not the sentence-initial transition include the character following the shift key. However, we found strong evidence for an increased hesitation probability when the before-sentence interkey interval included the character following the shift key. Notably, the hesitation probability approached ceiling when the interkey interval following the shift key was added. In other words, the mixture model identified almost all before-sentence transitions as being hesitant as they -- in the majority of cases -- included two keystrokes while interkey intervals for all other transition locations -- and therefore the majority of the data -- did not. 